Ensemble Stability of Biological Networks

### 2014-09-23 paper.tex
add todo
clarify linear stability analysis

In order to discuss how different parts of the system influence each other, we shall generalize our definition to include open systems.  We do this by having some of our variables be control variables whose values are given as specified functions of time rather than determined by the dynamics.  Given several such open systems, we may combine them to produce a larger system by setting the control variables of a subsystem equal to the dynamical variables of another system.  For details and formalization in terms of operads see \cite{Vagner2014}.  Taking this notion to its logical extreme, one can dissect a system into a collection of one open system for each dynamical variable.  However, for most purposes, such a decomposition would be uninteresting and what we instead want is a decomposition into relatively self-contained modules.  In the section on connected components, we shall propose a way of accomplishing this based upon the topology of the system graph.

@@

Another question one can ask is that of robustness, which measures the
probability that, if some property holds for a set of parameters, it will
continue to hold if we make a random perturbation about those parameter values.
To define robustness, we will introduce, in addition to the probability
distribution $\mu$ described above, a family of probability distributions
$\mu' \colon S \times \mathcal{B}(S) \to [0,1]$.  Then, given a binary random
variable $q$, we define its robustness as the following conditional proability:
\begin{align}\label{eq:robustness}
  R (q) =
  \frac{\int_S d\mu(x) \int_S d\mu'(x,y) \mathbf{1}_q(x) \mathbf{1}_q(y)}
  {\int_S d\mu(x) \mathbf{1}_q(x)}
\end{align}

\section{Stability analysis of biological networks}
The class of dynamical quantities on which we shall focus in this investigation involve stability of equilibria.  Suppose that $\vec x$ is a point on our phase space which depends upon the parameters.  Linearization of the $F_i$ about any fixed point $\vec{x}^0$ such that $F_i(\vec{x}^0)=0$ leads to the system
\begin{equation}\label{eq:lineardynsys}
\frac{d\vec{y}(t)}{dt} = A \vec{y}(t),
\end{equation}
where $\vec{y} = \vec{x} - \vec{x}^0$ and the $n \times n$-matrix $A$ has components
$$
a_{ij} = \left. \frac{\partial F_i}{\partial x_j} \right|_{\vec{x} = \vec{x}^0}.
$$
To each dynamical system having Jacobian matrix $A$ at some fixed point $\vec{x}^0$ we can associate a directed graph $G_A$ given by an adjacency matrix $\adj(G_A)$ where
 \begin{displaymath}
   \adj(G_A)_{ij} = \left\{
     \begin{array}{lr}
       1, & a_{ij} \neq 0\\
       0, & a_{ij} = 0
     \end{array}
   \right.
\end{displaymath}
The graphs $G_X$ and $G_A$ are equivalent because the condition $F_i$ independent of $x_j$ definitive of $\adj(G_X)$ corresponds precisely to $\frac{\partial F_i}{\partial x_j}=0$.

@@

To determine whether a randomly selected dynamical system evaluated at a random critical point, given by a matrix such as $A$, is stable, it is sufficient to check the above condition. Integrating over the region of the parameter space defining $A$ enables the determination of the probability of stability to perturbations in state $\vec{x}$ over an ensemble of dynamical systems.

% Another form of stability, referred to as structural stability \cite{Smale1967}, requires the determination of whether or not a given dynamical system that is determined to be stable remains stable under a perturbation to one of its defining parameters. This can be formalized as the conditional probability distribution
% $$
% P(A' \, \textrm{stable}\, \big| \, A \, \textrm{stable})
% $$
% where $A'$ represents a matrix derived from some perturbation applied to the matrix $A$ given above. In the simple case where $A \in \mathbb{R}^{n \times n}$ the perturbation which involves sampling uniformly over the $a_{ij}$ defining $A$ the value of a particular $a_{ij}$ from a given probability distribution.
Suppose that $M$ is an $n \times n$ matrix with real coefficients and connectivity $\connectivity$. We will consider the case in which each of the parameters defining the matrix is sampled from the uniform distribution $\mathcal{U}(-1,1)$ so that the parameter space corresponds to the $\connectivity$-dimensional hypercube, $H^\connectivity$, of edge length $r=2$, centered about the origin, and having volume $r^\connectivity$.  We want to know the probability that, if we resample the elements $k_1, k_2, \ldots k_m$ of this matrix, the resulting matrix will correspond to a stable matrix.  This is a special case of \ref{eq:robustness}:
\begin{equation}\label{eq:condprobgen}
 R (\mathrm{stab}, k_1, \ldots k_m) =
  \frac{\int_{H^\connectivity} dx_1 \cdots dx_d \int_{H^m} {dx'}_{k_1} \cdots {dx'}_{k_m}\,
    \mathbf{1}_{stab} \mathbf{1}_{stab'}}
  {\int_{H^\connectivity} dx_1 \cdots dx_d  \, \mathbf{1}_{stab}(x_1 \cdots x_d)}
\end{equation}
where we made the abbreviations
\begin{align*}
\mathbf{1}_{stab} &= \mathbf{1}_{stab}(x_1 \cdots x_{k_1} \cdots x_{k_m} \cdots x_\connectivity) \\
\mathbf{1}_{stab'} &= \mathbf{1}_{stab'}(x_1 \cdots {x'}_{k_1} \cdots {x'}_{k_m}  \cdots x_d)
\end{align*}
i.e. $\mathbf{1}_{stab'}$ corresponds to replacing $x_{k_1}, \ldots x_{k_m}$ with their primed counterparts.

@@

We note from the definition that the mapping from $G$ to $\hier(G)$ is many-to-one as exemplified in \ref{fig:scc}B and C.
\todo{explain how robustness in \ref{eq:robustness} is symmetric under permutations of the SCCs in $G$ or equivalently of the nodes of $\hier(G)$.}

@@

Using these observations, we may express the robustness of a graph in terms of the robustnesses of its strongly connected components.
If the components $C_i$ of our graph $G$ has $v_i$ vertices connectivity $\connectivity_i$ and there are $\ell_{ij}$ links between components $i$ and $j$, then the total robustness to resampling a random entry is
\begin{align}\label{eq:sccrobustness}
R(G) = \frac{\sum_{i,j \in \hier(G)} \ell_{ij} + \sum_{i=1}^n \connectivity_i R(C_i)}{\sum_{i,j \in \hier(G)} \ell_{ij} + \sum_{i=1}^n \connectivity_i}.
\end{align}
For instance, if our graph is the one in \ref{fig:scc}B or C, then we have two connected components, one with two nodes, and one with one node.  From \ref{tab:structstabmat}, we know that the graph with two nodes has probability $0.25$ of being stable and and robustness $0.62$.  The graph with one node corresponds to a $1 \times 1$ matrix, so we have probability $0.5$ of stability and robustness $0.5$.  Thus, the probability of our 3-node graph being stable is $0.5 \times 0.25 = 0.125$ and its robustness is
\[
R(G) = \frac{2 + 0.5 + 4 \times 0.62}{2 + 1 + 4} = 0.714,
\]
which agrees with the value computed in \ref{tab:structstabmat}.

@@

@@

\section{Tables}
\input{tex/tablenotation}
\input{tex/table2x2stab}
\input{tex/tablenographs}

>>>>

In order to discuss how different parts of the system influence each
other, we shall generalize our definition to include open systems.  We
do this by having some of our variables be control variables whose
values are given as specified functions of time rather than determined
by the dynamics.  Given several such open systems, we may combine them
to produce a larger system by setting the control variables of a
subsystem equal to the dynamical variables of another system.  For
details and a discussion of the operad structure of composition of
open systems see \cite{Vagner2014}.  Taking this notion to its logical
extreme, one can dissect a system into a collection of one open system
for each dynamical variable.  However, for most purposes, such a
decomposition would be uninteresting and what we instead want is a
decomposition into relatively self-contained modules.  In the section
on connected components, we shall propose a way of accomplishing this
based upon the topology of the system graph.

@@

A quality which is of interest to us is robustness or structural
stability \cite{Smale1967}, which requires the determination of
whether or not a given dynamical system that is determined to be
stable remains stable under a perturbation to one or more of its
defining parameters.  We will quantify this as the probability that,
if some property holds for a set of parameters, it will continue to
hold if we make a random perturbation about those parameter values.
To do so, we will introduce, in addition to the probability
distribution $\mu$ described above, a family of probability
distributions $\mu' \colon S \times \mathcal{B}(S) \to [0,1]$ such
that $\mu'(x,y)$ encodes the conditional probability that a system
with a parameter value $x$ will have its parameter values changed to
$y$ under a perturbation.  For our biological models, we will be
interested in large perturbations rather than the small or
infinitessimal perturbations usually considered in the theory of
dynamical systems.  Then, given a binary random variable $q$,
we define its robustness as the following conditional proability:
\begin{align}\label{eq:robustness}
  R (q) =
  \frac{\int_S d\mu(x) \int_S d\mu'(x,y) \mathbf{1}_q(x) \mathbf{1}_q(y)}
  {\int_S d\mu(x) \mathbf{1}_q(x)}
\end{align}

\section{Stability analysis of biological networks}
The class of dynamical quantities on which we shall focus in this
investigation involve stability of equilibria.  Suppose that $\vec x$
is a point on our phase space which depends upon the parameters.  Then
we may take $q$ to be a binary random variable which describes whether
or not $\vec x$ is a fixed point of the system (i.e. $q$ is true if and only if
$F_i(\vec{x}^0)=0$ for all $i$) and perform a probabilistic analysis
of the sort discussed above.

Furthermore, if it turns out that $\vec x$ is indeed a fixed point, we
may proceed to ask whether it is a dynamically stable fixed point.
Intuitively, dynamic stability means that, if one chooses the initial
conditions sufficiently close to the fixeed point, the solution will
stay close to the fixed point.  Physically, this is important because,
if a fixed point ${\vec x}_0$ is unstable, we have zero proability of
observing the solution ${\vec x}(t) = {\vec x}_0$.

To determine stability, we linearize the equations of motion about the
fixed point $\vec{x}^0$:
\begin{equation}\label{eq:lineardynsys}
\frac{d\vec{y}(t)}{dt} = A \vec{y}(t),
\end{equation}
where $\vec{y} = \vec{x} - \vec{x}^0$ and the $n \times n$-matrix $A$ has components
$$
a_{ij} = \left. \frac{\partial F_i}{\partial x_j} \right|_{\vec{x} = \vec{x}^0}.
$$
To each dynamical system having Jacobian matrix $A$ at some fixed point $\vec{x}^0$ we can associate a directed graph $G_A$ given by an adjacency matrix $\adj(G_A)$ where
 \begin{displaymath}
   \adj(G_A)_{ij} = \left\{
     \begin{array}{lr}
       1, & a_{ij} \neq 0\\
       0, & a_{ij} = 0
     \end{array}
   \right.
\end{displaymath}
The graphs $G_X$ and $G_A$ are equivalent because the condition $F_i$ independent of $x_j$ definitive of $\adj(G_X)$ corresponds precisely to $\frac{\partial F_i}{\partial x_j}=0$.

@@

Even though the determination of stability only requires examining
linearized equations of motion, it is worth noting that knowing the
stability of a fixed point can yield information about the non-linear
dynamics of a system.  For instance, in two dimensions,
Poincare-Bendixson theory implies that every limit cycle must encircle
at least one fixed point.  Furthermore, in the case where the cycle
contains a single fixed point (such as in the Volterra-Lotke model of
conflicting populations), the stability of the fixed point will be
determined by whether the limit cycle is attractive or repulsive.

%% To determine whether a randomly selected dynamical system evaluated at
%% a random critical point, given by a matrix such as $A$, is stable, it
%% is sufficient to check the above condition. Integrating over the
%% region of the parameter space defining $A$ enables the determination
%% of the probability of stability to perturbations in state $\vec{x}$
%% over an ensemble of dynamical systems.

% Another form of stability, referred to as structural
% stability \cite{Smale1967}, requires the determination of whether or
% not a given dynamical system that is determined to be stable remains
% stable under a perturbation to one of its defining parameters. This
% can be formalized as the conditional probability distribution
% $$
% P(A' \, \textrm{stable}\, \big| \, A \, \textrm{stable})
% $$
% where $A'$ represents a matrix derived from some perturbation
% applied to the matrix $A$ given above. In the simple case where
% $A \in \mathbb{R}^{n \times n}$ the perturbation which involves
% sampling uniformly over the $a_{ij}$ defining $A$ the value of a
% particular $a_{ij}$ from a given probability distribution.

Suppose that our Jacobian $A$ is an $n \times n$ matrix with real
coefficients and connectivity $\connectivity$ and denote the
proposition ``$A$ is stable'' as $\mathrm{stab}(A)$.  Then we may
compute the robustness of this proposition using \ref{eq:robustness}
once we have specified the proability densities $\mu$ and $\mu'$.  In
general, these will depend upon the parameters of the non-linear
system.  For the puprpose of the current investigation, we shall make
the ansatz that these are such that the distribution of entries of $A$
is (at least approximately) the uniform distribution
$\mathcal{U}(-1,1)$ on the $\connectivity$-dimensional hypercube,
$H^\connectivity$, of edge length $r=2$, centered about the origin.
To model the perturbations, we shall resample the elements $k_1,
k_2, \ldots k_m$ of this matrix from the same uniform distribution.
Under these assumptions, the expression for robustness becomes
of \ref{eq:robustness}:
\begin{equation}\label{eq:condprobgen}
 R (\mathrm{stab}, k_1, \ldots k_m) =
  \frac{\int_{H^\connectivity} dx_1 \cdots dx_d \int_{H^m} {dx'}_{k_1} \cdots {dx'}_{k_m}\,
    \mathbf{1}_{stab} \mathbf{1}_{stab'}}
  {\int_{H^\connectivity} dx_1 \cdots dx_d  \, \mathbf{1}_{stab}(x_1 \cdots x_d)},
\end{equation}
where we made the abbreviations
\begin{align*}
\mathbf{1}_{stab} &= \mathbf{1}_{stab}(x_1 \cdots x_{k_1} \cdots x_{k_m} \cdots x_\connectivity) \\
\mathbf{1}_{stab'} &= \mathbf{1}_{stab'}(x_1 \cdots {x'}_{k_1} \cdots {x'}_{k_m}  \cdots x_d)
\end{align*}
i.e. $\mathbf{1}_{stab'}$ corresponds to replacing $x_{k_1}, \ldots x_{k_m}$ with their primed counterparts.

@@

This map $\hier$ is many-to-one and so there is a large class of
operations which leaves $\hier(G)$ invariant for a given graph $G$.
For instance, we may interchange the positions of the strongly
connected components relative to each other.  Leaving the components
fixed, we may move links between nodes in a component or betwwen
components, or even add or delete links.  Some examples of these
operations are illustrated in \ref{fig:scc}B and C.  As we shall see
later, these operations leave invariant key dynamical quantities, so
we may regard them as a symmetry groupoid of our system.

@@

Using these observations, we may express the robustness of a graph in
terms of the robustnesses of its strongly connected components.  If
the components $C_i$ of our graph $G$ has $v_i$ vertices connectivity
$\connectivity_i$ and there are $\ell_{ij}$ links between components
$i$ and $j$, then there are a total of
$\sum_{i,j \in \hier(G)} \ell_{ij} + \sum_{i=1}^n \connectivity_i$
links in $G$.  If we resample a randomly chosen link in a component
$C_i$, the probability that the system will remain stable after
resamling is $R(\hbox{avg stab} C_i, 1)$.  If we resample a link
between components $C_i$ and $C_j$, the system will remain stable with
probability 1.  Hence, the probability that the system will remain
stable upon resampling a random link is the weighted average of these
probabilities:
\begin{align}\label{eq:sccrobustness}
R(\hbox{avg stab} G, 1) =
\frac{\sum_{i,j \in \hier(G)} \ell_{ij} +
      \sum_{i=1}^n \connectivity_i R(C_i)}
     {\sum_{i,j \in \hier(G)} \ell_{ij} +
      \sum_{i=1}^n \connectivity_i}.
\end{align}
For instance, if our graph is the one in \ref{fig:scc}B or C, then we have two connected components, one with two nodes, and one with one node.  From \ref{tab:structstabmat}, we know that the graph with two nodes has probability $0.25$ of being stable and and robustness $0.62$.  The graph with one node corresponds to a $1 \times 1$ matrix, so we have probability $0.5$ of stability and robustness $0.5$.  Thus, the probability of our 3-node graph being stable is $0.5 \times 0.25 = 0.125$ and its robustness is
\[
R(G) = \frac{2 + 0.5 + 4 \times 0.62}{2 + 1 + 4} = 0.714,
\]
which agrees with the value computed in \ref{tab:structstabmat}.

@@

This argument also works when we resample more than one entry,
although the notation becomes more complicated.  Suppose that we
resample $m$ nodes.  Then, given a partition $m = \sum_{i=0}^n m_i$,
there are ${m \choose m_0, m_1, \ldots. m_n}$ ways of choosing $m_i$
links from $C_i$ and $m_0$ links between strongly connected
components.  Hence, our weighted average becomes
\begin{equation}
R(\hbox{avg stab} C_i, m) =
\frac{sum_{m_0, m_1, \ldots m_n} {m \choose m_0, m_1, \ldots. m_n}
      m_0 + \sum_{i=1}^n m_i C_i}
     {}
\end{equation}

@@

\section{Tables}
\todo{Fill out notation table}
\input{tex/tablenotation}
\input{tex/table2x2stab}
\input{tex/tablenographs}
